---
title: "Darwin Core mapping"
subtitle: "For: De Vlaamse Waterweg target species list"
author:
- Damiano Oldoni
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

<!-- For examples, see https://github.com/trias-project/checklist-recipe/wiki/Examples -->

# Setup 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```

Install required libraries (only if the libraries have not been installed before):

```{r}
installed <- rownames(installed.packages())
required <- c( "readr", "dplyr", "magrittr", "here", "rgbif", "digest")
if (!all(required %in% installed)) {
  install.packages(required[!required %in% installed])
}
```

Load libraries:

```{r message = FALSE}
library(readr)          # To work with text files
library(dplyr)          # To manipulate tabular data
library(magrittr)       # To use %<>% pipes
library(here)           # To find files
library(rgbif)          # To use GBIF services
library(digest)         # To generate hashes
```

# Read source data

Create a data frame `input_data` from the source data:

```{r}
input_data <- readr::read_csv(file = here("data", "raw", "input_taxa.csv")) 
```

Preview data:

```{r}
input_data %>% head(n = 5)
```

# Process source data

## Scientific names

Use the [GBIF nameparser](https://www.gbif.org/tools/name-parser) to retrieve nomenclatural information for the scientific names in the checklist:

```{r}
parsed_names <- input_data %>%
  distinct(scientific_name) %>%
  pull() %>%
  rgbif::parsenames() # An rgbif function
```

Show scientific names with nomenclatural issues, i.e. not of `type = SCIENTIFIC` or that could not be fully parsed. Note: these are not necessarily incorrect.

```{r}
parsed_names %>%
  select(scientificname, type, parsed, parsedpartially, rankmarker) %>%
  filter(!(type == "SCIENTIFIC" & parsed == "TRUE" & parsedpartially == "FALSE"))
```

## Taxon ranks

The nameparser function also provides information about the rank of the taxon (in `rankmarker`). Here we join this information with our checklist. Cleaning these ranks will done in the Taxon Core mapping:

```{r}
input_data %<>% left_join(
  parsed_names %>%
  select(scientificname, rankmarker),
  by = c("scientific_name" = "scientificname"))
```

## Taxon IDs

To link taxa with information in the extension(s), each taxon needs a unique and relatively stable `taxonID`. Here we create one in the form of `dataset_shortname:taxon:hash`, where `hash` is unique code based on scientific name and kingdom (that will remain the same as long as scientific name and kingdom remain the same):

```{r}
vdigest <- Vectorize(digest) # Vectorize digest function to work with vectors
input_data %<>% mutate(taxon_id = paste(
  "dvw-target-list",
  "taxon",
  vdigest(paste(scientific_name, "Plantae"), algo = "md5"),
  sep = ":"
))
```

# Taxon core

We start by creating a copy of the input data:

```{r}
taxon <- input_data
```

## Term mapping

Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).

Start with record-level terms which contain metadata about the dataset (which is generally the same for all records).

### language

```{r}
taxon %<>% mutate(dwc_language = "en")
```

### license

```{r}
taxon %<>% mutate(dwc_license = "http://creativecommons.org/publicdomain/zero/1.0/")
```

### rightsHolder

```{r}
taxon %<>% mutate(dwc_rightsHolder = "INBO") # e.g. "INBO"
```

### datasetID

```{r}
taxon %<>% mutate(dwc_datasetID = "") # to be added after first publication 
```

### institutionCode

```{r}
taxon %<>% mutate(dwc_institutionCode = "INBO")
```

### datasetName

```{r}
taxon %<>% mutate(dwc_datasetName = "De Vlaamse Waterweg target species list")
```

The following terms contain information about the taxon:

### taxonID

```{r}
taxon %<>% mutate(dwc_taxonID = taxon_id)
```

### scientificName

```{r}
taxon %<>% mutate(dwc_scientificName = scientific_name)
```

### kingdom

```{r}
taxon %<>% mutate(dwc_kingdom = "Plantae")
```

### taxonRank

Inspect values:

```{r}
taxon %>%
  group_by(rankmarker) %>%
  count()
```

Map values by recoding to the [GBIF rank vocabulary](http://rs.gbif.org/vocabulary/gbif/rank_2015-04-24.xml):

```{r}
taxon %<>% mutate(dwc_taxonRank = dplyr::case_match(rankmarker,
  "sp." ~ "species"
))
```

Inspect mapped values: 

```{r}
taxon %>%
  group_by(rankmarker, dwc_taxonRank) %>%
  count()
```

## Post-processing

Only keep the Darwin Core columns:

```{r}
taxon %<>% select(starts_with("dwc_"))
```

Drop the `dwc_` prefix:

```{r}
colnames(taxon) <- gsub(
  x = colnames(taxon), 
  pattern = "dwc_", 
  replacement = ""
)
```

Preview data:

```{r}
taxon %>% head()
```

Save to CSV:

```{r}
write_csv(taxon, here("data", "processed", "taxon.csv"), na = "")
```
